---
title: "Intro to R: Day 1"
author: "Isaac Vock"
date: "2025-1-21"
categories: [stats]
format: 
  html:
    toc: true
engine: knitr
editor: 
  markdown: 
    wrap: 72
---

This worksheet will walk you through some basic concepts in R. I would
suggest copying code shown here into an R script and running it yourself
so that you can play around with the presented examples.

## Pre-requisite knowledge for `simple_calc()`

### Math in R

The simplest use case of R is using it to do math:

```{r}
1 + 1 # Addition
10 - 1.5 # Subtraction
2 * 3 # Multiplication
10.12 / 17.99 # Division
5 ^ 2 # Exponentiation
```

### Numeric variables in R

You can store numbers in "variables". This is like a special box in your
computer's memory labeled with a name (like `my_number`). When you put a
number into this box (for example, `10`), we say you have assigned the
value 10 to the variable `my_number`.

In R, you'd do this by writing:

```{r}
my_number <- 10
my_number
```

::: {.callout-tip collapse="true"}
## Seeing the value of variables

Typing and executing `print(my_number)` or just `my_number` will print
out the value of the variable to your console.
:::

Here is what's happening in this code:

1.  `my_number` is the label on the box in memory.
2.  `<-` is like an arrow pointing from the value to the box, meaning
    "put this value in that box".
3.  \`10 is the actual number you are storing.

You can then do math like this just like with regular numbers:

```{r}
my_number * 2
my_number ^ 3
my_number - 4
```

`my_number` does not change value in any of the above lines. To change
the value of `my_number`, you would have to assign it the new value with
`<-` again:

```{r}
my_number # my_number is 10
my_number <- 1001 # my_number is now 1001
my_number # Confirm new value of my_number
```

### Strings in R

You can store more than numbers in variables. For example, you can store
text, which is referred to as a "string":

```{r}
my_string <- "Hello"
my_string2 <- 'Bye'

my_string
my_string2
```

You tell R that you are storing text by wrapping that text in `""` or
`''`.

::: {.callout-tip collapse="true"}
## Useful string tools (aka functions)

Below are some useful tools that R provide you to work with strings.
These are called functions, a concept discussed later.

1.  `paste(..., sep = " ")`: `paste()` allows you to stitch together
    multiple strings, with a chosen separator text between strings
    (`sep` argument). Having no separator (`sep = ""`) is identical to
    using a different function `paste0()`:

```{r}
string1 <- "Hello"
string2 <- "friend."
string3 <- "It's been too long"

paste(string1, string2)
paste(string1, string2, sep = "")
paste0(string1, string2)
paste(string1, string2, string3)
paste(string1, string2, collapse = "_")
```

2.  `nchar()`: This will give you the number of individual characters in
    your text string:

```{r}
string1 <- "Hello"
nchar(string1)
```

3.  `gsub(pattern, replacement, x)`: This allows you to look for the
    string `pattern` in the query string `x`, and replace it with the
    string `replacement`:

```{r}
text <- "Hello, Hello, Hello!"
gsub("Hello", "Hi", text)
```

4.  `grepl(pattern, x)`: This is similar to `gsub()` but just searches
    for string `pattern` in string `x` and spits out `TRUE` if it finds
    it

```{r}
text <- "Hello, Hello, Hello!"
grepl("Hello", text)
```

There is a whole R package called
[stringr](https://stringr.tidyverse.org/) devoted to making working with
strings in R easier and more intuitive, so you might want to look into
that as well!
:::

### Booleans in R

Another thing that is commonly stored in variables is logical values
(TRUE or FALSE), otherwise known as "booleans":

```{r}
my_bool <- TRUE
my_bool2 <- FALSE

my_bool
my_bool2
```

You can do a sort of math with booleans, referred to as "boolean logic".
This takes as input two (in the case of AND and OR) or one (in the case
of NOT) boolean variables and outputs a new boolean. The most common
examples are:

**AND (`&`)**

-   Both of the booleans must be TRUE for the output to be TRUE:

```{r}
TRUE & TRUE # This is TRUE
TRUE & FALSE # This is FALSE
FALSE & TRUE # This is FALSE
FALSE & FALSE # This is FALSE
```

**OR (`|`)**

-   At least one of the booleans must be TRUE for the output of this to
    be TRUE

```{r}
TRUE | TRUE # This is TRUE
TRUE | FALSE # This is TRUE
FALSE | TRUE # This is TRUE
FALSE | FALSE # This is FALSE
```

**NOT (`!`)**

-   Unlike AND and OR, this takes a single boolean value as input
-   This reverses the value of the boolean:

```{r}
!TRUE # This is FALSE
!FALSE # This is TRUE
```

Finally, you can compare the value of two variables to see if they are
the same. If the are `variable_1 == variable_2` will return `TRUE`,
otherwise it will return `FALSE`:

```{r}
"Hello" == "Hello" # TRUE
"Hi" == "Bye" # FALSE
1 == 1 # TRUE

my_number <- 1
my_number2 <- 2
my_number == my_number2
```

### Functions in R

A function in R is like a "recipe" for a mini "machine" that does one
specific job. You give it some inputs (called *arguments*), it follows
the steps you've defined, and then it gives you a result.

Functions help you organize your code so you can reuse it instead of
writing the same steps again and again. Here is a simple example:

```{r}
# Function name: my_function
# Arguments: x and y
# Output: x + y
my_function <- function(x, y){
  
  # 1. Inside the curly braces, write the steps of what you will do with x and y
  
  # We will add x and y
  result <- x + y
  
  # 2. Tell the function what to output (i.e., its "return value")
  return(result)
  
}
```

-   `my_function` is the name of the function (like a label on the mini
    machine).
-   `function(x,y) { ... }` says "I am creating a function that expects
    two inputs, called `x` and `y`.
-   Inside the `{ ... }`, you can write as much code as you want; this
    is the instructions for what you want the function to do with the
    inputs
-   `return(result)` sends the output of the function back to you.

After creating `my_function`, you can *call* it (computer science lingo
meaning "use the function") by typing:

```{r}
my_function(3,5)
my_new_number <- my_function(2, 2)
my_new_number
```

Sometimes, you want one (or more) of your function's inputs to have a
"fallback" value if the user doesn't supply one. That's where **default
arguments** come in. For example:

```{r}
my_new_function <- function(x, y = 10){

  result <- x + y
  
  return(result)
  
}
```

`my_new_function` now only needs you to supply x. You can supply `x` and
`y`, but if you don't supply `y`, it will give `y` a default value of 10
by default:

```{r}
my_new_function(x = 1)
my_new_function(x = 2, y = 20)
my_new_function(2, 20) # Will fill arguments in order, so x = 2 and y = 20 here
```

::: {.callout-tip collapse="true"}
## Advanced function advice

### Tip 1: argument with small set of possible values

Sometimes, one of the arguments of your function may have a set number
of possible values that you intend for a user to input. You can specify
this as such:

```{r}
my_options <- function(a, b, greeting = c("Hi", "Bye", "Huh?")){
  
  # Check to make sure the user supplied one of the valid options
  greeting <- match.arg(greeting)
  
  print(greeting)
  
  result <- a + b
  
  return(result)
  
}

# Uses first option by default
my_options(2, 2)

my_options(2, 2, "Huh?")

```

### Tip 2: Catching errors

In all of our examples so far, we have assumed that the user has
supplied a particular kind of data for each argument. Mostly, we have
assumed that numbers are being passed to many of our example functions,
numbers that we can add. What if they messed up though and passed a
string, for example? We can catch this and throw an error message:

```{r}

my_valuecheck <- function(a, b){
  
  # check if a is a number
  stopifnot(is.numeric(a))
  
  # check if b is a number, but with a slightly different strategy
  # if-statements are discussed more later.
  if(!is.numeric(b)){
    
    stop("b isn't a number")
    
  }
  
  result <- a + b
  
  return(result)
}
```

This function will work as normal if `a` and `b` are numbers, but will
throw informative error messages if not. You will also get an error in
the old version of this function that didn't have the `stopifnot()`
lines, but this error might be far more cryptic and hard to understand.
You will also get a different error depending on what is wrong about `a`
and/or `b`, further confusing you or other users of this function.
:::

### Control flow (if-else statements)

An if-else statement is one of the most common ways to control the flow
of a program. It lets your code make decisions based on whether a
condition is `TRUE` or `FALSE`.

-   `if` checks if something is `TRUE`
-   `else` covers what happens if it is not `TRUE`
-   You can add extra steps in between using `else if` to handle
    different possible conditions

The basic structure looks like:

```{r}
#| eval: false

if (condition1){
  # This code runs if 'condition1' is TRUE
}else if(condition2){
  # This code runs if 'condition2' is TRUE
}else{
  # This code runs if both 'condition1' and 'condition2' are FALSE
}
```

Think of this code as asking a set of questions:

-   If condition1 is `TRUE`, do something.
-   Else if condition2 is `TRUE`, do something else
-   Else, if neither condition1 nor condition2 are `TRUE`, do a default
    thing.

A real example might look like:

```{r}
x <- 5

if(x > 3){
  print("x is greater than 3")
}else if(x < 5){
  print("x is between 3 and 5")
}else{
  print("x is greater than or equal to 5")
}
```

Conditions in R must evaluate to a single `TRUE` or `FALSE`. Common ways
to form conditions are comparison operators:

-   `==`: Check if two things are equal (e.g., `a == b`). `a` and `b`
    can be numbers, strings, booleans, etc.
-   `!=`: Check if two things are not equal 1 `<`, `>`, `<=`, `>=`: Less
    than, greater than, less than or equal to, or greater than or equal
    to, respectively.

Here is an example of how you might use control flow in a function:

```{r}
greetUser <- function(user_input){
  
  # Check if user_input equals "Hello"
  if (user_input == "Hello"){
    return("Hi there! Nice to meet you.")
  }else if(user_input == "Goodbye"){
    return("See you later! Take care.")
  }else{
    return("I'm not sure how to respond to that...")
  }
  
}

greetUser("Hello")
greetUser("Comment allez-vous?")
```

## Pre-requisite knowledge for `vector_calc()`

### Vectors

In R, a vector is a container that holds multiple values of the **same
data type** (such as numbers, strings, or booleans). You can think of it
like a row of boxes, each containing a value of the same kind.

You can create a vector with the `c()` function (short for "combine" or
"concatenate"). Here are a few example:

```{r}
# A numeric vector
numbers <- c(10, 20, 30, 40)

# A character (string) vector
words <- c("cat", "dog", "bird")

# A boolean vector
bools <- c(TRUE, FALSE, TRUE)
```

Often, you will want to access specific elements or sets of elements of
a vector. To do this, you can use square brackets `[ ]`:

```{r}
# Get the first element of 'numbers'
numbers[1] # 10

# Get the second element of 'words'
words[2]

# Get multiple elements at once
numbers[c(1, 3)] # This gives the 1st and 3rd elements: c(10, 30)

# Exclude specific elements
bools[-1] # This gives everything but the 1st element
```

You can also change values of specific elements:

```{r}
# See what 'numbers' is now
numbers

# Change a value
numbers[2] <- 99

# Check 'numbers' now
numbers
```

Sometimes, it will be useful to check what kind of data is in a vector.
This can be done with the `class()` function:

```{r}
class(numbers) # "numeric"
class(words) # "character"
class(bools) # "logical" (another word for boolean)
```

You can also check the value with functions like `is.numeric()`.
`is.character()`, or `is.logical()`:

```{r}
is.numeric(numbers) # TRUE
is.numeric(words) # FALSE

is.character(numbers) # FALSE
is.logical(numbers) # FALSE

is.character(words) # TRUE
is.logical(bools) # TRUE
```

::: {.callout-tip collapse="true"}
## Useful vector functions

Below are some useful functions that allow you to create vectors or
lookup some information about a vector:

1.  `length(v)`: returns the number of elements in the vector v:

```{r}
length(c(1, 2, 3))
```

2.  `seq(from, to, length.out)` or `seq(from, to, by)`: Creates a vector
    starting from the number `from` (default value of 1), to the number
    `to` (default value of 1). If you set `length.out`, then you will
    get a vector of `length.out` elements. If you set `by`, then you
    specify the distance between adjacent elements:

```{r}
seq(from = 1, to = 5, length.out = 5)
seq(from = 1, to = 5, by = 1)
```

3.  `rep(x, times)`: Creates a vector containing the value `x` repeated
    `times` times:

```{r}
rep(x = 1, times = 10)
```

4.  `start:end`: Same as `seq(from = start, to = end, by = 1)`:

```{r}
1:10
0.5:2.5
```
:::

### Loops

A loop is a way to tell R to "do something multiple times". This unlocks
one of the powerful aspects of computers: their ability to do multiple
things quickly.

There are two commonly used types of loops: **for** loops and **while**
loops.

A **for** loop in R iterates (or "loops") over each element of a vector
and does something with it. For example, if we want to print every
element of a numeric vector:

```{r}
numbers <- c(10, 20, 30, 40)

# Loop over the values
for(value in numbers){
  print(value)
}


# Loop over the vector 1 to the length of the vector
for(i in 1:length(numbers)){
  
  print(numbers[i])
 
}

# Fancier alternative to the above code
for(i in seq_along(numbers)){
  
  print(numbers[i])
  
}

```

What's happening here?

1.  `for (value in numbers)` means "go through each element of `numbers`
    and temporarily call that element `value`.
    `for(i in 1:length(numbers)` creates a vector (`1:length(numbers)`)
    which is a vector of whole numbers from 1 to the length of the
    vector numbers. Each of these whole numbers is then temporarily
    called `i`. `seq_along(numbers)` does pretty much the same things as
    `1:length(numbers)`.
2.  `print(value)` means we display the current `value` on the screen.
3.  R will do this until it has gone through all elements in `numbers`.

A **while** loop keeps going as long as some condition is `TRUE`.
Suppose we want to keep adding numbers from a vector until the total sum
exceeds 50:

```{r}
numbers <- c(10, 20, 30, 40, 50)
total <- 0 # Start total at 0
i <- 1 # Start index at 1

while(i <= length(numbers) & total <= 50){
  
  # Add to total
  total <- total + numbers[i]
  
  # Track which element we are on
  i <- i + 1
  
}

print(total)
print(i)
```

What's going on here?

1.  `while(i <= length(numbers) && total <= 50)` - The loop will
    continue running **while** two conditions are both `TRUE`:

-   We haven't reached the end of the vector (`i <= length(numbers)`)
    and
-   The `total` hasn't exceeded 50 (`total <= 50`).

1.  Inside the loop, we add the `i`-th element of `numbers` to `total`.
2.  We then move `i` to the next element by adding 1.
3.  As soon as one of the conditions in 1. become `FALSE`, the loop
    stops.

## Pre-requisite knowledge for `calc_df_stats()`

### Navigating directories and file paths

When you work in R, you'll often deal with files (like CSV files) that
sit in folders (directories) on your computer. To load these files into
R so that you can work with and analyze them, you need to tell R where
they are. In addition, it is important to know where you are while
working in R.

When I say "know where you are", I am referring to your "working
directory". When you open up Rstudio, there is some folder on your
computer that R will call your "working directory". You can see what
this folder is at any time with `getwd()`:

```{r}
# You get to see what my working directory path is
getwd()
```

If you want to change this directory, you can switch to a new directory
with `setwd("/path/to/new/working/directory")`.

You can specify a file path to `setwd()` in one of two ways:

1.  As an **absolute path**. This means specifying *exactly* where a
    folder is on your computer (like "C:/Users/YourName/Documents/" on
    Windows or "/Users/YourName/Documents/ on Macs).
2.  As a **relative path**. This means telling R how to get from the
    current working directory to the folder. For example, if you are in
    the directory "Documents" and want to move to a folder called "data"
    inside of "Documents, you could run `setwd("data")`, assuming
    "Documents" is your current working directory.

### Reading a file with readr

The [readr](https://readr.tidyverse.org/) package (part of the tidyverse
collection of packages), provides user-friendly functions for reading in
data. For example, you can read a csv file like so:

```{r}
#| eval: false
my_data <- read_csv("path/to/mydata.csv")
```

-   `read_csv("path/to/mydata.csv")` reads the CSV file located at the
    specified path (either a relative or absolute path) and creates a
    data frame (more on those soon).
-   We're storing that data frame in a variable called `my_data`.

### Data Frames

A data frame is a table-like structure with rows and columns, commonly
used for storign datasets in R. Each **column** is usually a vector of a
particular type (numeric, character, boolean, etc.), and all columns
have the **same length**.

To create a data frame you can run code like this:

```{r}
ages <- c(30, 25, 35)

# You can either specify the vector directly
# or provide the name of a vector you previously created
people_df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = ages,
  Score = c(100, 95, 90)
)

people_df
```

Here are some ways you can interact with the data inside of a data
frame:

1.  You can grab an entire column with `$` or
    `[[<col name as string>]]`:

```{r}
# This will give you a vector
people_df$Name

# This will be the same vector
people_df[["Name"]]
```

2.  You can grab an entire column with `[ , <column number>]`:

```{r}
# This will give you a data frame with one column
people_df[, 1]

# This will give you a data frame with multiple columns
people_df[,c(1, 2)]
```

3.  You can get all columns in a specific row with `[<row number>, ]`:

```{r}
# This will give you a data frame with one row
people_df[1,]

# This will give you a data frame with multiple rows
people_df[c(1, 3), ]
```

### Lists

A list is like a container in R that can hold a mix of different types
of items, like a data frame. Lists are more flexible though, and can
hold things of different sizes. A list can hold:

-   A numeric vector
-   A string vector
-   A single number
-   An entire data frame
-   Another list

All at once!

Here is how you can create a list:

```{r}
my_list <- list(
  name = "Alice",
  age = 30,
  scores = c(100, 95, 90),
  is_student = FALSE,
  df = data.frame(a = c(1, 2, 3), b = c("a", "b", "c"))
)
```

To access elements of a list, you can:

1.  Use the `$` operator (if the elements have names):

```{r}
my_list$name

my_list$scores
```

2.  The `[[ ]]` operator with the element's name (if it has one), or its
    position:

```{r}
my_list[["name"]]

my_list[[1]]
```

::: {.callout-tip collapse="true"}
## Fancy looping over lists (and data frames)

Often, you will want to go element by element of a list and do something
with each element. In addition, data frame columns are equivalent to
elements of a list (actually, under the hood, a data frame is just a
list that forces the list elements to be the same size). You could write
a for loop, but there are popular alternatives that can make your code
cleaner and easier to read. R has a version of these, but the R package
`purrr` has improved versions of these that I prefer.

1.  `map()`: takes a single list as input

```{r}
library(purrr)

numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(10, 20, 30, 40, 50)
)

# Outputs a list, one element for original element of the list
map(numbers, function(x) sum(x))

# Outputs a vector numbers, one element per original list element
# Also using an alternative notation
map_dbl(numbers, ~ sum(.x))

```

2.  `map2()`: takes two lists as input

```{r}

numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(10, 20, 30, 40, 50)
)

numbers2 <- list(
  c(-1, -2, -3),
  c(12, 13),
  c(2, 4, 6, 8, 10, 12)
)

# Outputs a list, one element for original element of the list
map2(numbers, numbers2, ~ sum(.x) + sum(.y))

# Outputs a vector numbers, one element per original list element
map2_dbl(numbers, numbers2, function(x, y) sum(x) + sum(y))


```

3.  `pmap()` allows you to provide a named list of inputs:

```{r}
# A list of vectors
lists_of_inputs <- list(
  a = c(1, 3, 5),
  b = c(2, 4, 6),
  c = c(10, 20, 30)
)

pmap(lists_of_inputs, function(a, b, c) a + b + c)
pmap_dbl(lists_of_inputs, function(a, b, c) a + b + c)
```
:::
